---
title: "Johns Hopkins Covid-19 Data Analysis for DTSA-5301"
author: "Daniel South"
date: "2025-03-04"
output: pdf_document
---

# Johns Hopkins Covid-19 Data Analysis

COVID-19 (SARS-CoV-2) caused millions of deaths as it spread around the globe. Johns Hopkins tracked confirmed cases and deaths in the United States and worldwide in an effort to better understand how it was being transmitted and where it would be likely to cause issues in the near future.

This report reviews time series tracking data from Johns Hopkins and plots the number of cases over time in various regions.

```{r libs, include=FALSE}
library(tidyverse)
library(lubridate)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Import, Clean Up, and Prepare Data Sets

## Import Data Files

```{r importData}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv"
                )
urls <- str_c(url_in, file_names)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```

\newpage

## Pivot, Drop, and Rename Columns

```{r tidyData}

tidy_gbl_cases = global_cases %>%
  pivot_longer(cols = -c("Province/State", "Country/Region", "Lat", "Long"),
               names_to="the_date",
               values_to="num_cases") %>%
  select(-c("Lat", "Long"))

tidy_gbl_cases = tidy_gbl_cases %>%
  rename('Country_Region' = 'Country/Region', 'Province_State' = 'Province/State') %>%
  mutate(the_date = mdy(the_date))



tidy_gbl_deaths = global_deaths %>%
  pivot_longer(cols = -c("Province/State", "Country/Region", "Lat", "Long"),
               names_to="the_date",
               values_to="num_deaths") %>%
  select(-c("Lat", "Long"))

tidy_gbl_deaths = tidy_gbl_deaths %>%
  rename('Country_Region' = 'Country/Region', 'Province_State' = 'Province/State') %>%
  mutate(the_date = mdy(the_date))


tidy_us_cases = US_cases %>%
  pivot_longer(cols = -c("UID", "iso2", "iso3", "code3", "FIPS", "Admin2", "Combined_Key", "Province_State", "Country_Region", "Lat", "Long_"),
               names_to="the_date",
               values_to="num_cases") %>%
  select(-c("Lat", "Long_", "iso2", "iso3", "code3", "FIPS"))

tidy_us_cases = tidy_us_cases %>%
  rename('County' = 'Admin2') %>%
  mutate(the_date = mdy(the_date))


tidy_us_deaths = US_deaths %>%
  pivot_longer(cols = -c("UID", "FIPS", "Combined_Key", "Admin2",  "iso2", "iso3", "code3",  "Province_State", "Country_Region", "Population", "Lat", "Long_"),
               names_to="the_date",
               values_to="num_deaths") %>%
  select(-c("Lat", "Long_", "iso2", "iso3", "code3", "FIPS"))

tidy_us_deaths = tidy_us_deaths %>%
  rename('County' = 'Admin2') %>%
  mutate(the_date = mdy(the_date))

```

\newpage

## Joins and Transformations

```{r joinAndTransform}
global_data = tidy_gbl_cases %>%
  full_join(tidy_gbl_deaths)

us_data = tidy_us_cases %>%
  full_join(tidy_us_deaths)

global_data = global_data %>% filter(num_cases > 0)

us_data = us_data %>% filter(num_cases > 0)
```

```{r addColsToGLobal}
# Add Combined_Key and Population to the Global Data Set

global_data = global_data %>% 
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid = read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global_data2 = global_data %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, the_date, num_cases, num_deaths, Population, Combined_Key)
```

\newpage

## Summaries of Prepared Data Sets

```{r summaryOfDataSets}
summary(us_data)
summary(global_data2)
```



\newpage

# Part 2 - Data Visualization

## Summarize US Data

```{r summarizeData}
# Summarize US Data

US_by_state = us_data %>%
  group_by(Province_State, Country_Region, County, the_date) %>%
  summarize(num_cases = sum(num_cases), num_deaths = sum(num_deaths), Population = max(Population)) %>%
  mutate(deaths_per_mill = num_deaths * 1000000/Population) %>%
  select(Province_State, Country_Region, the_date, num_cases, num_deaths, deaths_per_mill, Population, County) %>%
  ungroup()



US_totals = US_by_state %>%
  group_by(Country_Region, the_date) %>%
  summarize(num_cases = sum(num_cases), num_deaths = sum(num_deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = num_deaths * 1000000/Population) %>%
  select(Country_Region, the_date, num_cases, num_deaths, deaths_per_mill, Population) %>%
  ungroup()

```

\newpage

## Derive Columns for New Cases and Deaths

```{r newCases}
US_by_state = US_by_state %>%
  mutate(new_cases = num_cases - lag(num_cases),
         new_deaths = num_deaths - lag(num_deaths))

US_totals = US_totals %>%
  mutate(new_cases = num_cases - lag(num_cases),
         new_deaths = num_deaths - lag(num_deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))
```


\newpage

## Visualize US Data

```{r visUSData}
US_totals %>%
  ggplot(aes(x = the_date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in the US", y = NULL)
```

\newpage

## Visualize NY State Data

```{r}
state_name = "New York"

US_by_state %>%
  filter(Province_State == state_name) %>%
  filter(num_cases > 0) %>%
  ggplot(aes(x = the_date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state_name), y = NULL)
```

\newpage

# Part 3 - Data Analysis



```{r dataAnalysis}
US_state_totals = US_by_state %>%
  group_by(Province_State) %>%
  summarize(num_deaths = max(num_deaths), num_cases = max(num_cases),
            population=max(Population),
            cases_per_thou = 1000 * num_cases / population,
            deaths_per_thou = 1000 * num_deaths / population
            ) %>%
  filter(num_cases > 0, population > 0)




```

## US States with the Highest Death Rates

```{r highestRates}
US_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```

\newpage

## US States with the Lowest Death Rates

```{r lowestRates}
US_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

```

\newpage

# Part 4 - Data Modeling

If we fit the US state totals to a linear model, we can use visualization to determine whether the model can predict cases accurately.

## Summary of the Linear Model applied to US State Totals

```{r dataModeling}
lin_mod = lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(lin_mod)
```

```{r linearPredictions}
predict(lin_mod)

state_predicitons = US_state_totals %>% mutate(pred = predict(lin_mod))

state_predicitons %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color="blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color="red")
```


\newpage

# Part 5 - Conclusions

### Potential Biases

Efforts to mitigate the spread of Covid-19 became controversial. Businesses closures led to financial hardships. Supply shortages led to difficult conditions in hospitals. Lockdowns led to feelings of isolation. Infection rates rose and fell in unpredictable patterns. People have strong opinions to this day regarding public health measures that were deployed during the pandemic.

It's conceivable that cases were under-reported in some countries and states. Politicians in some areas encouraged suppression of infection counts to encourage a rapid end to mitigation efforts.

### Predictions

The linear model would not have served as an effective predictor of Covid-19 cases. Additional data points would be needed in order to create a useful model.

